Environmental Factors Affecting Water Quality at Toronto Beaches

### 2024-09-27 paper/paper.qmd
update paper

library(tidyverse)
library(opendatatoronto)
library(kableExtra)
library(here)
library(tibble)
```

@@

# Load the clean data
cleaned_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

@@

# Data {#sec-data}
## Data Overview {#sec1-data} 
To investigate water quality and beach conditions in Toronto, I found the ‘Toronto Beach Observations’ dataset from the Open Data Toronto portal through the library opendatatoronto (@opendatatoronto). The dataset contains more than 18,000 observations of daily beach conditions and includes a number of various environmental variables, including air temperature, water temperature, wind speed, and waterfowl population, which play a significant role in determining beach water safety. The analysis focuses on the key variables affecting water turbidity, which is a proxy for water pollution risk. All variables are explained and described in (Section A.[1](#appendix-a.1)) In my this report, I will focuses on 8 key variables: date, beach name, wind's speed, air temperature, wether rain or not, water temperature, waterfowl, water turbidity.

@@

## Cleaned Data {#sec2-data} 
During the data cleaning process, I took several important steps to ensure that the dataset was suitable for analysis. First, all rows containing missing values (NA) were removed. This decision was made to avoid any potential biases or inaccuracies that might result from incomplete data, as missing values could skew the results of the statistical analyses. Also, I removed several outliers. Some of the outliers such as the water temperature was up to 220 degree centigrade, which is a very unusual value. So I set reasonable range values for water temperature, air temperature, and wind speed, and removed any unreasonable values.

I cleaned and prepared a subset of the data as shown in @tbl-sum1 below. I showed the first and last 5 rows of the key variables in this table.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum1
#| fig-cap: "Displaying the first and last 5 rows of seven key variables"

# Generate table
kbl(sample_data, 
    col.names = c("Date", "Beach", "Wind (km/h)", "Air Temp (°C)", "Rain", "Water Temp (°C)", "Fowl Count", "Turbidity (NTU)"),
    ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center",
                font_size = 9) %>%
  column_spec(1, bold = TRUE, width = "6em") %>%  
  column_spec(2, width = "8em") %>%  
  column_spec(3:4, width = "5em") %>%  
  column_spec(5, width = "2em") %>%  
  column_spec(6, width = "6em") %>%
  column_spec(7:8, width = "3em")

```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum2
#| fig-cap: "Summary statistics table"

# Display the summary statistics table in a vertically aligned format
kbl(summary_stats, col.names = c("Variable", "Mean", "SD", "Min", "Median", "Max")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center", 
                font_size = 10) %>%
  row_spec(0, bold = TRUE)

```
@tbl-sum2 summaries the numerical data. The average wind speed was 13.3km/h with a maximum of 50 km/h, which stirs up sediments and may increase water turbidity. The average air temperature was 19.9°C, which directly affects the comfort of beachgoers and water quality, since warm air leads to higher water temperatures and increased microbial activity. Water temperatures averaged 15.6°C, reaching a maximum of 27°C during the summer months. this is often associated with bacterial growth, leading to health warnings. The number of waterfowl is another important factor, with an average of 32.3 waterfowl, but sometimes the number spikes to 450. The presence of waterfowl can bring faecal pollution into the water which increases the risk of bacterial contamination and water quality warnings.

##Data Visualization {#sec3-data} 
### Turbidity across different beach 
@fig-box1 provides a clear visualization of the distribution of turbidity across various beach locations. Bluffer's Beach Park, Kew Balmy Beach, and Cherry Beach appear to have the highest overall turbidity levels, with numerous outliers reaching above 50 NTU and some exceeding 100 NTU. Hanlan’s Point Beach, Ward’s Island Beach, Gibraltar Point Beach, and Centre Island Beach exhibit much lower turbidity levels overall. Other beaches are showing more variability than low-turbidity beaches but less than the higher-turbidity ones.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-box1
#| fig-cap: "Distribution of Turbidity by Beach"

# Boxplot of turbidity by beach
ggplot(key_variables, aes(x = reorder(beachName, turbidity, FUN = median), y = turbidity)) +
  geom_boxplot(fill = "skyblue", color = "black", outlier.color = "red", outlier.shape = 16) +
  labs(x = "Beach Name", y = "Turbidity (NTU)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

@@

### Turbidity Levels on Rainy vs. Non-Rainy Day
@fig-box2 shows the distribution of turbidity levels (measured in NTU) on rainy versus non-rainy days. On rainy days, the change in turbidity increases relatively significantly. The occurrence of multiple outliers above 50 NTU indicates that rainfall is largely responsible for the increase in turbidity. The larger interquartile range (IQR) indicates that turbidity values are more scattered on rainy days compared to non-rainy days.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-box2
#| fig-cap: "Turbidity Levels on Rainy vs. Non-Rainy Days"

@@

# Appendix {#appendix-a}
## Variable description{#appendix-a.1}
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum3
#| fig-cap: "Description of variables in the dataset"

data <- data.frame(
  column = c("_id", "dataCollectionDate", "beachName", "windSpeed", 
             "windDirection", "airTemp", "rain", "rainAmount", "waterTemp", 
             "waterFowl", "waveAction", "waterClarity", "turbidity"),

  description = c("Unique row identifier for Open Data database", 
                  "Date observations were collected", 
                  "Name of beach where observations were collected", 
                  "Wind speed measured in km/h", 
                  "Wind directions - East, North-East, North, North-West, West, South-West, South, South-East", 
                  "Air temperature in degrees Celsius", 
                  "Presence of rain - Yes, No", 
                  "Amount of rainfall in mm in the last 24 hours", 
                  "Water temperature in degrees Celsius", 
                  "Count of water fowl seen", 
                  "Observations on wave action - high, low, moderate, none", 
                  "Visual state of water's clarity - free form text", 
                  "Turbidity of water measured in Nephelometric Turbidity Units (NTUs)")
)
data$description <- str_wrap(data$description, width = 25)

kable(data, 
      align = "l", 
      col.names = c("Column", "Description"),
      escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = F, 
                position = "center", 
                font_size = 10)
```

>>>>

library(tidyverse)
library(opendatatoronto)
library(kableExtra)
library(here)
library(tibble)
library(dplyr)
library(knitr)
```

@@

# Load the clean data
cleaned_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

@@

# Data {#sec-data}
## Data Overview {#sec1-data} 
To investigate water quality and beach conditions in Toronto, I found the ‘Toronto Beach Observations’ dataset from the Open Data Toronto portal through the library opendatatoronto (@opendatatoronto). The dataset contains more than 18,000 observations of daily beach conditions and includes a number of various environmental variables, including air temperature, water temperature, wind speed, and waterfowl population, which play a significant role in determining beach water safety. The analysis focuses on the key variables affecting water turbidity, which is a proxy for water pollution risk. All key variables are explained and described in (Section A.[1](#appendix-a.1)) In my this report, I will focuses on 8 key variables: date, beach name, wind's speed, air temperature, wether rain or not, water temperature, waterfowl, water turbidity.

@@

I cleaned and prepared a subset of the data as shown in (Section A.[2](#appendix-a.2)). (Section A.[2](#appendix-a.2)) described the process of my data cleaning. I showed the first and last 5 rows of the key variables in @tbl-sum2. Also, I summarize the numerical data in @tbl-sum3.

## Data Visualization{#sec2-data}
### Turbidity across different beach 
@fig-box1 provides a visualization of the distribution of turbidity across various beach locations. Since this diagram looked squeezed, I used a logarithmic scale for to make trends more apparent @fig-box2. The new diagram looks to contrast very well. Bluffer's Beach Park, Kew Balmy Beach, and Cherry Beach appear to have the highest overall turbidity levels, with numerous outliers reaching above 50 NTU and some exceeding 100 NTU. Hanlan’s Point Beach, Ward’s Island Beach, Gibraltar Point Beach, and Centre Island Beach exhibit much lower turbidity levels overall. Other beaches are showing more variability than low-turbidity beaches but less than the higher-turbidity ones.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-box1
#| fig-cap: "Distribution of Turbidity by Beach"

# Boxplot of turbidity by beach
ggplot(key_variables, aes(x = reorder(beachName, turbidity, FUN = median), y = turbidity)) +
  geom_boxplot(fill = "cyan", color = "black", outlier.color = "red", outlier.shape = 16) +
  labs(x = "Beach Name", y = "Turbidity (NTU)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-box2
#| fig-cap: "Distribution of Turbidity by Beach (Log Scale)"

# Boxplot of turbidity by beach with log scale
ggplot(key_variables, aes(x = reorder(beachName, turbidity, FUN = median), y = turbidity)) + 
  geom_boxplot(fill = "cyan", color = "black", outlier.color = "red", outlier.shape = 16) +
  scale_y_log10() +  # Apply log scale transformation to y-axis
  labs(x = "Beach Name", y = "Turbidity (NTU, Log Scale)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

@@

### Turbidity Levels on Rainy vs. Non-Rainy Day
@fig-box2 shows the distribution of turbidity levels (measured in NTU) on rainy versus non-rainy days. On rainy days, the change in turbidity increases relatively significantly. The occurrence of multiple outliers above 50 NTU indicates that rainfall is largely responsible for the increase in turbidity. The larger interquartile range (IQR) indicates that turbidity values are more scattered on rainy days compared to non-rainy days.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-box3
#| fig-cap: "Turbidity Levels on Rainy vs. Non-Rainy Days"

@@

# Appendix {#appendix-a}
## Variable description{#appendix-a.1}
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum1
#| fig-cap: "Variable description"

data <- data.frame(
  Column = c("dataCollectionDate", "beachName", "windSpeed", "airTemp", "rain", "waterTemp", "waterFowl", "turbidity"),
  Description = c(
    "Date observations were collected",
    "Name of beach where observations were collected",
    "Wind speed measured in km/h",
    "Air temperature in degrees Celsius",
    "Presence of rain - Yes, No",
    "Water temperature in degrees Celsius",
    "Count of water fowl seen",
    "Turbidity of water measured in Nephelometric Turbidity Units (NTUs)"
  )
)

data |>
  kable(col.names = c("Column", "Description"), align = c("l", "l"), escape = FALSE, booktabs = TRUE) |>
  kable_styling(full_width = F, position = "center", font_size = 10) |>
  row_spec(0, bold = TRUE, color = "black", background = "lightgray") |> 
  column_spec(1, bold = TRUE, width = "15em") |> 
  column_spec(2, width = "30em") |> 
  footnote(general = "Note: Description of variables used in the Toronto beach water quality dataset.", general_title = " ")

## Cleaned Data {#appendix-a.2}
During the data cleaning process, I took several important steps to ensure that the dataset was suitable for analysis. First, all rows containing missing values (NA) were removed. This decision was made to avoid any potential biases or inaccuracies that might result from incomplete data, as missing values could skew the results of the statistical analyses. Also, I removed several outliers. Some of the outliers such as the water temperature was up to 220 degree centigrade, which is a very unusual value. So I set reasonable range values for water temperature, air temperature, and wind speed, and removed any unreasonable values.

@tbl-sum2 summaries the numerical data. The average wind speed was 13.3km/h with a maximum of 50 km/h, which stirs up sediments and may increase water turbidity. The average air temperature was 19.9°C, which directly affects the comfort of beachgoers and water quality, since warm air leads to higher water temperatures and increased microbial activity. Water temperatures averaged 15.6°C, reaching a maximum of 27°C during the summer months. this is often associated with bacterial growth, leading to health warnings. The number of waterfowl is another important factor, with an average of 32.3 waterfowl, but sometimes the number spikes to 450. The presence of waterfowl can bring faecal pollution into the water which increases the risk of bacterial contamination and water quality warnings.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum2
#| fig-cap: "Displaying the first and last 5 rows of seven key variables"

# Create the formatted table using kable and kableExtra
sample_data %>%
  kable(
    col.names = c("Date", "Beach", "Wind (km/h)", "Air Temp (°C)", "Rain", "Water Temp (°C)", "Fowl Count", "Turbidity (NTU)"), # Column headers
    digits = 1,            # Number of decimal places for numerical columns
    booktabs = TRUE,       # Enable booktabs for improved styling
    align = c("l", "l", "c", "c", "c", "c", "r", "r"),  # Align columns (left, center, right)
    format.args = list(big.mark = ",")  # Use commas as thousand separators for large numbers
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), # Bootstrap styling options
    full_width = FALSE,   # Prevent the table from taking full width
    position = "center",  # Center align the table
    font_size = 8        # Set font size
  ) %>%
  row_spec(0, bold = TRUE, background = "lightgray", color = "black") %>%  # Style header row
  column_spec(1, width = "6em") %>%    # Set width and style for Date column
  column_spec(2, width = "8em") %>%                 # Set width for Beach column
  column_spec(3:4, width = "4em") %>%               # Set width for Wind and Air Temp columns
  column_spec(5, width = "3em") %>%                 # Set width for Rain column
  column_spec(6, width = "5em") %>%                 # Set width for Water Temp column
  column_spec(7:8, width = "5em")                   # Set width for Fowl Count and Turbidity columns

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-sum3
#| fig-cap: "Summary numerical variables"

summary_stats %>%
  kable(
    col.names = c("Variable", "Mean", "SD", "Min", "Median", "Max"), 
    booktabs = TRUE,  
    align = c("l", "c", "c", "c", "c", "c"),   
    format.args = list(big.mark = ",")  
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = F,   
    position = "center",   
    font_size = 10   
  ) %>%
  row_spec(0, bold = TRUE, background = "lightgray", color = "black") %>%  # Style header row
  footnote(general = "Summary statistics for Wind Speed, Air Temperature, Water Temperature, Fowl Count, and Turbidity.")
```
